Colonies - Compute Continuums across Platforms

### 2023-04-02 paper_2023/paper.tex
wip

To address the aforementioned challenges, the paper introduces a framework termed Colonies designed to facilitate the execution of computational workloads across a diverse range of platforms. Colonies is founded upon a loosely-coupled microservice architecture that beaks down complex workflows into composable functions. With the use of an HTTP protocol, these composable functions can be implemented in any computer language and be executed by independent executors deployed across various systems, e.g. cloud, edge, devices, or even in web browsers. Colonies orchestrates the execution and by using a zereo-trust security protocol, a collection of distributed executors can function as a single cohesive unit, thereby establishing a seamless compute continuum across multiple platforms.

@@

Taking an AI model into production requires substantial software engineering expertise. In contrast to traditional IT workloads, both the data and the model must be managed in addition to the software itself. As most models require regular re-training or re-calibration, it must be possible to update deployed models and software seamlessly without losing information or breaking the system. In many cases, there is a constant flow of data ingested into the system which must be managed even if some components are malfunctioning. This becomes even more challenging when nodes or parts of the underlying infrastructure become unavailable due to maintenance or failaure such as software updates, hardware replacements or simply missconfiguration.

@@

In reality, AI system requires integration of multiple systems. For instance, data may need to be captured from an IoT system or pulled from third-party database running on different domains than the compute cluster. With the emergence of edge computing, parts of a data pipeline may also run on edge servers to bring computations closer to data sources. Configuring and setting up such pipelines add even more complexity.

@@

This paper presents a framework called Colonies, which is built around a loosely-coupled microservices architecture that separates workflow definitions from implementation and deployment. The main objective is to create a tool where monolithic workflows can be broken down into independent and separated compute units that can be dynamically added or removed while executing workflows. These compute units can be implemented in any computer language and be deployed anywhere on the Internet.

\section{Related work}
A workflow is a series of tasks that need to be completed in a specific order. Workflows are often represented as directed acyclic graphs (DAGs), where nodes represent tasks and edges represent dependencies or data flow between tasks. Workflow management has been extensively studied in both academic and industrial settings, with numerous approaches proposed to address the challenges in this field.

Another popular approach is the use of workflow engines, which provide a framework for creating and executing workflows. Workflow engines can be integrated with other tools and systems to automate complex business processes. Examples of workflow engines include Apache Airflow, Camunda, and Zeebe.

Apache Airflow is a popular open-source workflow management system used to programmatically author, schedule, and monitor workflows.
Luigi is also an open-source workflow management system developed by Spotify. It is designed to handle long-running batch jobs and data pipelines.

Pinball is a workflow management system developed by Pinterest that allows users to define and execute workflows using Python scripts. It has a simple architecture, with support for DAGs and task dependencies. However, it lacks some of the advanced features of Airflow, such as task retries and monitoring.

TODO:

@@

Queues enable different software components to communicate indirectly by acting as a buffer between them. Queues makes it possible to decouple each executor and make them operate independently, e.g. an executor can be updated or replaced without affecting other executors. Queues also allow for asynchronous communication between executors, enabling them to process tasks at their own pace. This ensures that slower executors do not bottleneck faster ones, leading to a more efficient and scalable system. Most importantely, queues enable load balancing by distributing tasks among multiple executors, thus making it possible to parallelize workflow execution.

Queues can be implemented in different ways, and while message brokers are a common solution, Colonies has adopted an alternative strategy by leveraging a standard database and querying it for tasks to assign to different executors. One key advantage of this approach is that it enables fine-grained task assignments, making it possible to assign specific tasks to particular executors. For instance, an executor of the browser type can be limited to only executing tasks of the browser type. This level of granularity cannot easily be implemented using message brokers which generally do not offer introspection of queue, or provide the ability to pull specific messages out of the queue. Generarally, the only way to retrieve a specific message is to pull all messages from the queue, obtain the message, and then place all remaining messages back into the queue in the same order. In contrast, using a database a query can match any columns and be ordered using a timestamp to make it function as a queue.

\begin{equation}
    priority_{time}=priority \cdot -10^9 \cdot 60 \cdot 60 \cdot 24 + submission_{time}  
\end{equation}

@@

\subsection{Process tables}
Colonies provides a platform for executors to interact with each other by enabling them to submit function specifications to a Colonies server. Once submitted, other executors can connect to the server to receive execution assignments. When a function specification is submitted to the Colonies server, it is store as a process entry in a dedicated process table database. In this case, a process is analogous to a task, but contains additional contextual information about the execution, for example, execution status (waiting, running, successful, failed), assigned executor Id, submission time, execution time, deadlines, priority etc. It also contains input and argument values as well as return values from the function innovation.

The Colonies server acts as a job broker for executors, almost like an employment agency for people. When an executor connects to the Colonies server, the server hang the incomming HTTP connection\footnote{An alternative protocol is to use WebSockets or gPRC to communicate with the Colonies server.} until the executor is assigned a process, or until a connection timer expires. Note that the Colonies server does not connect to any executor. Rather, it is the responsibility of the executors to connect to the Colonies server. This enables executors to be deployed anywhere on the Internet, even behind firewalls, commercial telco networks, or in application running in web browsers.

Figure \ref{fig:process_assignment} illustrates an executor submitting a function specification that is later assigned to another executor. Executors have to specify to the Colonies server which functions they are capable of executing. The Colonies server makes sure that the conditions (i.e requirmenets) of a function specification matches the capability of an executor.

\begin{figure}[h]
	\centering
    \includegraphics[scale=0.35]{function_spec.png}
	\caption{Example of a function spec.}
	\label{fig:function_spec}
\end{figure}

Figure \ref{fig:function_spec} shows an example of a function specification. The \emph{maxexectime} attribute specifies the maximum time an executor may run a process (in this case, 100 seconds). Before a process is assigned to an executor, the Colonies server updates the process entry in the process table database and calculates a deadline for when the process must be finished. The server then regularly checks for any running processes that have exceeded their deadlines. If such process is detected, it is reset, allowing it to be re-assigned to other executors. This scheme is completely stateless, so if the Colonies server crashes, another Colonies server can take over and look for processes that have exceeded their deadline. Moreover, if an executor assigned to a process crashes, the process is simply re-executed by another executor.

Making it possible to specify maximum execution time is a simple but powerful mechanism. To scale up a system, more executors can simply be deployed. Scaling down, however, can be more challenging. One solution is to select a set of executors to be removed and then starv them out by denying them new process assignments. Another, simpler solution, is to immediately destroy the executors, or if possible notify them in advance so that they can gracefully unregister.

@@

Utilizing the \emph{maxexectime} failsafe mechanism not only enhances system reliability, but also provides an opportunity to apply of Chaos engineering \cite{chaos_engineering}. For example, deliberately terminating executors using a Chaos monkey. If executors are deployed on Kubernetes, Kubernetes will automatically redeploy terminated executors. The constant replacement of executor instances ensures that the system is capable of gracefully tolerating failures.

@@

The primary objective of all Colonies API requests is to either alter a state stored in the database or retrieve information from it. The Colonies framework is designed to be stateless, meaning that the Colonies server does not keep any information between requests. In other words, each request is handled independently, without relying on any information from previous requests.

One example of a request is the assign request, which allocates a waiting process from the database to an executor. Given the multi-threaded nature of the Colonies server, it is essential that the assign request is synchronized to ensure that only one thread at a time can modify the database and update the process table, thus preventing multiple executors from being assigned to the same process. As only the assigned executor can modify a running process and only one executor must be assigned to a process, synchronization is required for the assign request. It is worth noting that synchronization is not necessary for other requests, such as the submit request or the close request. As the submit request only add new entries to the process table, there is no race conditions. The close request set the output of the function innovation and updates the process state to either successful or failed in the process table. Since there exists only one assigned executor there is no need for synchronization.

@@

\subsubsection{Workflows}
TODO
\begin{figure}[h]
	\centering
    \includegraphics[scale=0.30]{workflow.png}
	\caption{cron management}
	\label{fig:fig3}
\end{figure}

>>>>

To address the aforementioned challenges, the paper introduces a framework termed Colonies which is available as open-source\footnote{https://github.com/colonyos/colonies} and designed to facilitate execution of computational workloads across a diverse range of platforms. Colonies is founded upon a loosely-coupled microservice architecture that beaks down complex workflows into composable functions. With the use of an HTTP protocol, these composable functions can be implemented in any computer language and be executed by independent executors deployed across various systems, e.g. cloud, edge, devices, or even in web browsers. Colonies orchestrates the execution and by using a zereo-trust security protocol, a collection of distributed executors can function as a single cohesive unit, thereby establishing seamless compute continuums across multiple platforms.

@@

Taking an AI model into production requires substantial software engineering expertise. In contrast to traditional IT workloads, both the data and the model must be managed in addition to the software itself. As most models require regular re-training or re-calibration, it must be possible to update deployed models and software seamlessly without losing information or breaking the system. In many cases, there is a constant flow of data ingested into the system which must be managed even in case of failures. This becomes even more challenging when nodes or parts of the underlying infrastructure become unavailable due to maintenance such as software updates, hardware replacements and sometimes missconfiguration failures.

@@

In reality, AI system requires integration of multiple systems. For instance, data need to be captured from an IoT system or pulled from third-party database running on different domains than the compute cluster. With the emergence of edge computing, parts of a data pipeline may also run on edge servers to bring computations closer to data sources. Configuring and setting up such pipelines add even more complexity.

@@

This paper presents a framework called Colonies, which is built around a loosely-coupled microservices architecture that separates workflow definitions from implementation and deployment. The main objective is to create a tool where monolithic workflows can be broken down into independent and separated compute units that can be dynamically added or removed while executing workflows. These compute units can be implemented in any computer language and be deployed anywhere on the Internet. The remainder of the paper describes the Colonies framework and how it can be used to create robust and scalable AI systems.

\section{Related work}
Workflow management has been extensively studied in both academic and industrial settings \cite{apache_airflow}, with numerous approaches proposed to address the challenges in this field. For instance, Apache Airflow \cite{apache_airflow} is a popular open-source workflow management system for handling data engineering pipelines. Like Colonies, Apache Airflow enables developers to create custom operators and executors that can be integrated with various systems. Additionally, Apache Airflow offers an HTTP API that allows for the development of clients and software development kits (SDKs) in multiple programming languages.

However, Apache Airflow is not designed around a queuing system. Instead, it must be integrated with a message broker, such as RabbitMQ \cite{rabbitmq} or Kafka \cite{apache_kafka}, to implement task queues, resulting in a more complex architecture than Colonies. Furthermore, Colonies is based on a distributed microservice architecture that makes it more suitable for DevOps software development. As Colonies is loosely coupled, executors can be implemented and dynamically deployed without reconfiguring the workflow engine, effectively transforming it into a grid computing platform.

When it comes to academic works, \cite{GarciaRepresa1740746} t

@@

Queues enable software services to communicate indirectly by acting as a buffer between them. Queues makes it possible to decouple each executor and make them operate independently, e.g. an executor can be updated or replaced without affecting other executors. Queues also allow for asynchronous communication between executors, enabling them to process tasks at their own pace. This ensures that slower executors do not bottleneck faster ones, leading to a more efficient and scalable system. Most importantely, queues enable load balancing by distributing tasks among multiple executors, thus making it possible to parallelize workflow execution.

Queues can be implemented in different ways, and while message brokers are a common solution, Colonies has adopted an alternative strategy by leveraging a standard database and querying it for tasks to assign to different executors. One key advantage of this approach is that it enables fine-grained task assignments, making it possible to assign specific tasks to particular executors. For instance, an executor of the browser type can be limited to only executing tasks of the browser type. This level of granularity cannot easily be implemented using message brokers which generally do not offer introspection of queue, or provide the ability to pull specific messages out of the queue. Generarally, the only way to retrieve a specific message is to pull all messages from the queue, obtain the message, and then place all remaining messages back into the queue in the same order. In contrast, a database can also function as a queue and a query can match any columns thus making it possible to assign specific executors to processes.

\begin{equation}
    \label{eq:pt}
    priority_{time}=priority \cdot -10^9 \cdot 60 \cdot 60 \cdot 24 + submission_{time}  
\end{equation}

Implementing queue functionality using a database can be achieved in SQL by utilizing the \emph{order by timestamp} and \emph{limit 1} clauses to select tasks according to their submission times. Upon submission of a task, a \emph{priority time} value is computed and stored in the database. The primary objective of the priority time is to modify the submission timestamp by subtracting a delta time calculated as a function a priority value. This enables prioritization of more critical tasks. Equation \ref{eq:pt} shows how the priority time is calculated.

@@

\subsection{Process tables}
Colonies provides a platform for executors to interact with each other by enabling them to submit function specifications to a Colonies server. Once submitted, other executors can connect to the server to receive execution assignments. When a function specification is submitted to the Colonies server, it is store as a process entry in a process table database. In this case, a process is similar to a task or a job, but also contains contextual information about the execution, for example, execution status (waiting, running, successful, failed), assigned executor ID, submission time, execution time, deadlines, priority etc. It also contains input and argument values as well as return values from the function innovation. In the remainder of the paper, we are going to use the term process instead of task or job.

The Colonies server acts as a job broker for executors, almost like an employment agency for people. When an executor connects to the Colonies server, the server hang the incomming HTTP connection\footnote{An alternative protocol is to use WebSockets or gPRC to communicate with the Colonies server.} until the executor is assigned a process, or until a connection timer expires. Note that the Colonies server does not connect to any executor. Rather, it is the responsibility of the executors to connect to the Colonies server. This enables executors to be deployed anywhere on the Internet, behind firewalls, commercial telco networks, or even in web browser-based applications.

Figure \ref{fig:process_assignment} illustrates an executor submitting a function specification that is later assigned to another executor. When registering, executors have to specify to the Colonies server which functions they are capable of executing. The Colonies server makes sure that the conditions (i.e requirmenets) of a function specification matches the capability of an executor. 
\begin{figure}[h]
	\centering
    \includegraphics[scale=0.35]{function_spec.png}
	\caption{Example of a function spec.}
	\label{fig:function_spec}
\end{figure}

\subsection{A stateless failsafe mechanism}
The primary objective of all Colonies API requests is to alter a state stored in the database or retrieve information from the database. The Colonies framework is designed to be stateless, meaning that the Colonies server does not keep any information between requests. In other words, each request is handled independently, without relying on any information from previous requests.

Figure \ref{fig:function_spec} shows an example of a function specification. The \emph{maxexectime} attribute specifies the maximum time an executor may run a process (in this case, 100 seconds). Before a process is assigned to an executor, the Colonies server updates the process entry in the process table database and calculates a deadline for when the process must be finished. The server then regularly checks for any running processes that have exceeded their deadlines. If such process is detected, it is reset, allowing it to be re-assigned to other executors.

Making it possible to specify maximum execution time is a simple but powerful mechanism. To scale up a system, more executors can simply be deployed. Scaling down, however, can be more challenging. One solution is to select a set of executors to be removed and then starv them out by denying them new process assignments. Another, simpler solution, is to immediately destroy the executors and use the \emph{maxexectime} failsafe mechanism move back failed processes to the queue.

@@

Utilizing the \emph{maxexectime} failsafe mechanism not only enhances system reliability, but also provides an opportunity to apply Chaos engineering \cite{chaos_engineering}. For example, a Chaos monkey can be used to deliberately terminate executors. If executors are deployed on Kubernetes, Kubernetes will then automatically redeploy terminated executors. The constant replacement of executor instances ensures that the system is capable of gracefully tolerating failures.

@@

The assign request allocates binds a process from the database to an executor. Given the multi-threaded nature of the Colonies server, it is essential that the assign request is synchronized to ensure that only one thread at a time can modify the database and update the process table, thus preventing multiple executors from being assigned to the same process. As only the assigned executor can modify a running process and only one executor must be assigned to a process, synchronization is required for the assign request. It is worth noting that synchronization is not necessary for other requests. For example, as the submit request only add new entries to the process table, there is no race conditions. The close request set the output of the function innovation and updates the process state to either successful or failed in the process table. Since there is only one assigned executor to a process there is no race conditions and consequently no need for synchronization.

@@

\subsubsection{Workflows}
A workflow is a series of tasks that need to be completed in a specific order. Workflows are often represented as directed acyclic graphs (DAGs), where nodes represent tasks and edges represent dependencies or data flow between tasks.

\begin{figure}[h]
	\centering
    \includegraphics[scale=0.30]{workflow.png}
	\caption{cron management}
	\label{fig:fig3}
\end{figure}
